---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "80%",
  fig.height = 3,
  fig.align = "center"
)

library(tidyverse)
theme_set(theme_bw())
theme_update(text = ggplot2::element_text(size = 15))
```

# PDSIR

<!-- badges: start -->
<!-- badges: end -->

The `PDSIR` R package implements an efficient data augmentation MCMC (DA-MCMC) algorithm for exact Bayesian inference under the semi-Markov stochastic susceptible-infectious-removed (SIR) model, given discretely observed counts of infections.
The novelty of this DA-MCMC algorithm is the *joint* update of the high-dimensional latent data.
In a Metropolis-Hastings step, the latent data are jointly proposed from a surrogate process carefully designed to closely resemble the target process and from which we can efficiently generate epidemics consistent with the observed data. This yields a MCMC algorithm that explores the high-dimensional latent space efficiently, mixes significantly better than single-site samplers, and scales to outbreaks with thousands of infections.

The package contains data from the 2013-2015 outbreak of Ebola in Western Africa to illustrate the use of the algorithm on real data.

## Installation

You can install the development version of PDSIR from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("rmorsomme/PDSIR")
```

## Illustration on synthetic data

We start by generating artificial data from a semi-Markov SIR process with Weibull-distributed infection periods. We stop the process at time `t_end=4`, when the outbreak is not completely over.

```{r simulate-data}

library(PDSIR)

# Setup
set.seed(1)
S0 <- 150  # initial number of susceptible individuals
I0 <- 5    # initial number of infectious individuals

t_end <- 4 # # duration of observation period

iota_dist <- "weibull" # distribution of the infection periods
theta <- list(
  R0 = 2,               # basic reproduction number
  lambda = 1, shape = 2 # parameters of the Weibull distribution for the infection periods
  )  

theta <- complete_theta(theta, iota_dist, S0) # add the infection rate parameter, beta, used in the MCMC (and the average infection period, gamma)

# Simulate artificial data
SIR <- simulate_SEM(S0, I0, t_end, theta, iota_dist)

# Trajectories of compartments
draw_trajectories(SIR, t_end)
```

The observed data consist the number of infections in pre-specified intervals. here, we consider `K=10` intervals of equal length.

```{r observed-data}
K <- 10 # number of observation intervals
Y <- observed_data(SIR, K)

print(Y$ts ) # endpoints of the intervals
print(Y$I_k) # number of infections per interval
```

We run the DA-MCMC algorithm for `N=50,000` iterations using the (default) weakly informative prior $\beta \sim G(0.01, 1)$ and $\lambda \sim Ga(1,1)$ independently. The entire latent are updated each iteration (`rho=1`, by default).
The sampler is fast and achieves a healthy acceptance rate in the Metropolis-Hastings step for the latent data.

```{r, cache=TRUE}
out <- run_DAMCMC(Y, N = 5e4, iota_dist = iota_dist, theta_0 = theta)

print(out$run_time)    # run time in seconds
print(out$rate_accept) # acceptance rate in the Metropolis-Hastings step for the latent data
```

The traceplot for $R_0$ indicates that the chain mixes well. The true value is $R_0=2$ (red dotted line).

```{r synthetic-traceplot, echo=FALSE, message=FALSE}
library(tidyverse)

THETA <- out$theta %>%
      data.table::rbindlist() %>%
      tibble() %>%
      mutate(
        Iteration = 1:nrow(.),
#        `Expected Infection Period` = lambda^{-1/shape}*gamma(1+1/shape) # expectation of weibull distribution
        ) 
THETA %>%
  ggplot(aes(x = Iteration, y = R0)) +
  geom_line() +
  geom_hline(yintercept = theta$R0, col = "red", linewidth = 1.5, linetype= 2) + 
  theme(
#      axis.text.x = ggplot2::element_text(size = 25)
      ) + 
  labs(y = expression(R[0]))
```


## Illustration on the 2013-2015 outbreak of Ebola in Western Africa

a total of `r sum(ebola_gueckedou$I_k)` infections in the prefecture Gueckedou, Guinea.

```{r ebola, echo=FALSE}
tibble(
  time = ebola_gueckedou$dates,
  I_k  = ebola_gueckedou$I_k
  ) %>%
  ggplot(aes(time, I_k)) +
  geom_col() +
  labs(x = "time (weeks)", y = "Counts of infections")
```

```{r, cache=TRUE}
# set up
rm(list=ls())
set.seed(0)

S0 <- 292e3 # https://en.wikipedia.org/wiki/Prefectures_of_Guinea
I0 <- 5
t_end <- max(ebola_gueckedou$ts)

ebola_gueckedou$S0 <- S0
ebola_gueckedou$I0 <- I0
ebola_gueckedou$t_end <- t_end

iota_dist <- "weibull"
theta_0 <- list(R0 = 1, lambda = 0.01, shape = 2)  

theta_0 <- complete_theta(theta_0, iota_dist, S0)

N <- 1e6
thin <- 1e2

# run MCMC
out <- run_DAMCMC(
    ebola_gueckedou, N = N, rho = 0.1, iota_dist = iota_dist, thin = thin, theta_0 = theta_0
)


print(out$run_time)
print(out$rate_accept)
```

```{r ebola-traceplot, echo=FALSE, message=FALSE}
THETA <- out$theta %>%
      data.table::rbindlist() %>%
      tibble() %>%
      mutate(
        Iteration = 1:nrow(.),
        EIP = lambda^{-1/shape}*gamma(1+1/shape) # expectation of weibull distribution
        )

THETA %>%
  ggplot(aes(x = EIP)) +
  #geom_histogram(aes(y = .data$..count../sum(.data$..count..))) +
  geom_histogram(aes(y = after_stat(density))) +
  theme(
#      axis.text.x = ggplot2::element_text(size = 25)
      ) +
  labs(x = "Expected infection period", y = "Proportion")
```

```{r ebola-histogram, echo=FALSE}
THETA %>%
  ggplot(aes(x = Iteration, y = EIP)) +
  geom_line() +
  theme(
#      axis.text.x = ggplot2::element_text(size = 25)
      ) + 
  labs(y = "Expected infection period")
```


You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>.

This package contains the code used in the paper "Uniformly Ergodic Data-Augmented MCMC for Fitting the General Stochastic Epidemic Model to Incidence Data" by R. Morsomme and J. Xu available on ArXiv. 
We use it to fit a semi-Markov susceptible-infectious-removed model to the 2013-2015 outbreak of Ebola Haemorrhagic Fever in Gu\'eck\'edou, Guinea.
