---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.height = 3
)
```

# PDSIR

<!-- badges: start -->
<!-- badges: end -->

The `PDSIR` R package implements an efficient data augmentation MCMC (DA-MCMC) algorithm for exact Bayesian inference under the semi-Markov stochastic susceptible-infectious-removed (SIR) model, given discretely observed counts of infections.
The novelty of this DA-MCMC algorithm is the *joint* update of the high-dimensional latent data.
In a Metropolis-Hastings step, the latent data are jointly proposed from a surrogate process carefully designed to closely resemble the target process and from which we can efficiently generate epidemics consistent with the observed data. This yields a MCMC algorithm that explores the high-dimensional latent space efficiently, mixes significantly better than single-site samplers, and scales to outbreaks with thousands of infections.


## Installation

You can install the development version of PDSIR from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("rmorsomme/PDSIR")
```

## Illustration of synthetic data

We employ the DA-MCMC on artifical data; semi-Marko model

This package contains the code used in the paper "Uniformly Ergodic Data-Augmented MCMC for Fitting the General Stochastic Epidemic Model to Incidence Data" by R. Morsomme and J. Xu available on ArXiv. 
We use it to fit a semi-Markov susceptible-infectious-removed model to the 2013-2015 outbreak of Ebola Haemorrhagic Fever in Gu\'eck\'edou, Guinea.

This is a basic example which shows you how to solve a common problem:

```{r synthetic-data}
library(PDSIR)

# Setup
set.seed(0)
S0 <- 500  # initial number of susceptible individuals
I0 <- 5    # initial number of infectious individuals

t_end <- 6 # # duration of observation period

iota_dist <- "weibull" # distribution of the infection periods
theta <- list(
  R0 = 2,               # basic reproduction number
  lambda = 1, shape = 2 # parameters of the Weibull distribution for the infection periods
  )  

theta <- complete_theta(theta, iota_dist, S0) # add the infection rate parameter, beta, and the average infection period, gamma

# Simulate artificial data
SIR <- simulate_SEM(S0, I0, t_end, theta, iota_dist)

# Trajectories of compartments
draw_trajectories(SIR, t_end)
```

Observed data

```{r observed-data}
K <- 10 # number of intervals
Y <- observed_data(SIR, K)

print(Y$ts ) # endpoints of intervals
print(Y$T_k) # number of infections per interval
```

## Illustration on the 2013-2015 outbreak of Ebola in Western Africa

You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>.

